% 1. Tell the TOC to start tracking sections again
\addtocontents{toc}{\protect\setcounter{tocdepth}{2}}

% 2. Style the TOC header
\etocsettocstyle{\subsection*{Contents of Appendix}}{}

% 3. Print the standard global TOC (which is now filtered)
\tableofcontents 
\clearpage


\section{Proof of the Unified Risk Bound (Theorem~\ref{thm:unified_bound})}
\label{app:detailed_proofs}

We provide the complete proof of Theorem~\ref{thm:unified_bound}, which the main text states for general $W \in \mathcal{O}(d)$ (trace family $\Omega_W$, Eq.~\ref{eq:omega_def}). The proof proceeds through five self-contained steps: risk decomposition via triangle inequality, Lipschitz analysis of cross-entropy with respect to features, the alignment-residual decomposition (whose Procrustes-optimal specialization $W = W_N$ yields the nuclear form $\Omega_N$ via SVD), second-moment-adjusted head bound, and final assembly. The identity-alignment specialization $W = I$ (used throughout Sec.~\ref{sec:experiments} and the regularizer of Sec.~\ref{subsec:shape_reg}) and the $W = W_N$ specialization (used in the Sec.~\ref{subsec:ablation} ablation) follow as corollaries of the same proof.

\subsection{Relation to Classical Procrustes Shape Metrics}
\label{app:shape_metrics_relation}

The left-hand side of Eq.~(\ref{eq:exact_equality}) at $W = W_N$ is the squared Procrustes size-and-shape distance $d_1^2(Z_T, Z_P)$ studied in the generalized shape metrics framework~\cite{williams2021generalized} and its recent decodability analysis~\cite{harvey2024what}. Expanding $d_1^2$ reveals that the nuclear norm $\|Z_T^\top Z_P\|_*$ appears inside any Procrustes-based similarity; accordingly, our contribution is \emph{not} the Procrustes distance itself. Rather, it is (a)~the explicit split of this residual into the scale term $(\rho_T-\rho_P)^2$ and the shape term $2\rho_T\rho_P(1-\Omega_N)$, which the shape-metric literature treats jointly as a single descriptive dissimilarity, and (b)~its lifting to a functional-risk bound (Theorem~\ref{thm:unified_bound}). Sections~\ref{subsec:quantization} and~\ref{subsec:forgetting} show that the two arms of this split have qualitatively different practical meanings and dominate the risk gap under different LLM lifecycle settings.


\subsection{Step 1: Risk Decomposition via Triangle Inequality}
\label{app:risk_decomp}

\paragraph{Setup.}
Recall that the risk of model $M$ is $\mathcal{R}_M = \mathbb{E}_{(x,y)}[\ell(\phi_M(x) \cdot H_M, y)]$, where $\ell$ is the cross-entropy loss (Eq.~\ref{eq:ce_def}). For any orthogonal alignment $W \in \mathcal{O}(d)$, introduce the hybrid risk:
\begin{equation}
\mathcal{R}_{P \to T} := \mathbb{E}_{(x,y)}\!\Big[\ell\!\big(\phi_P(x)\, W \cdot H_T,\; y\big)\Big].
\end{equation}
This evaluates the target's head $H_T$ on proxy features aligned by $W$, serving as a bridge between $\mathcal{R}_T$ and $\mathcal{R}_P$. The proof below holds for any $W \in \mathcal{O}(d)$; the two specializations relevant to the paper are $W = I$ (identity alignment, trace form $\Omega$, used throughout the main text) and $W = W_N := \arg\min_W \|Z_T - Z_P W\|_F^2$ (Procrustes-optimal, nuclear form $\Omega_N$, Sec.~\ref{subsec:ablation} ablation).

\paragraph{Decomposition.}
By the triangle inequality on absolute values:
\begin{equation}
\label{eq:app_triangle}
|\mathcal{R}_T - \mathcal{R}_P| \;\le\; \underbrace{\big|\mathcal{R}_T - \mathcal{R}_{P \to T}\big|}_{\text{Feature Alignment Error } \delta} \;+\; \underbrace{\big|\mathcal{R}_{P \to T} - \mathcal{R}_P\big|}_{\text{Head Discrepancy } \gamma}.
\end{equation}
$\delta$ measures how features differ through the same head $H_T$; $\gamma$ measures how heads differ on the same aligned features.


\subsection{Step 2: Bounding the Feature Alignment Error $\delta$}
\label{app:kfeat}

\paragraph{Lipschitz Property of Cross-Entropy with Respect to Features.}
We bound $\delta$ by showing that the cross-entropy loss is Lipschitz continuous in the feature vector $z$. Define the end-to-end composition $g_T(z, y) := \ell(z \cdot H_T, y)$, treating the classification head $H_T$ and the cross-entropy loss as a single map from features to scalar loss. We bound $g_T$ via direct gradient analysis rather than the Lipschitz composition rule (loss-Lipschitz $\sqrt{2}$ times head operator norm $\|H_T\|_2$): the composition rule is agnostic to the softmax output and gives the spectral constant $\sqrt{2}\|H_T\|_2$ (Remark below), whereas the gradient $\nabla_z g_T = H_T(\hat{p} - e_y)$ exposes the predicted distribution $\hat{p}$ and lets us derive a constant set by pairwise token-embedding distances instead. Our goal is to find the tightest constant $K_{\mathrm{feat}}$ such that:
\begin{equation}
|g_T(z_1, y) - g_T(z_2, y)| \le K_{\mathrm{feat}} \|z_1 - z_2\|_2, \quad \forall z_1, z_2, y.
\end{equation}

\paragraph{Computing the Gradient.}
The logit vector is $v = z \cdot H_T \in \mathbb{R}^V$, and the cross-entropy loss is $\ell(v, y) = -v_y + \log \sum_{j=1}^{V} e^{v_j}$. The gradient with respect to the logits is:
\begin{equation}
\frac{\partial \ell}{\partial v} = \hat{p} - e_y,
\end{equation}
where $\hat{p} = \mathrm{softmax}(v) \in \mathbb{R}^V$ and $e_y$ is the one-hot vector at index $y$. Applying the chain rule to obtain the gradient with respect to features:
\begin{equation}
\nabla_z \ell = H_T (\hat{p} - e_y) = \sum_{j=1}^{V} (\hat{p}_j - \delta_{jy})\, h_{T,j},
\end{equation}
where $h_{T,j}$ denotes the $j$-th column of $H_T$ (the embedding vector for token $j$).

\paragraph{Simplex Polarization.}
We now derive a tight bound on $\|\nabla_z \ell\|_2$ by exploiting the probability simplex constraint $\sum_j \hat{p}_j = 1$. Expanding the gradient for true class $y$:
\begin{equation}
\nabla_z \ell = (\hat{p}_y - 1)\, h_{T,y} + \sum_{j \neq y} \hat{p}_j\, h_{T,j}.
\end{equation}
Since $1 - \hat{p}_y = \sum_{j \neq y} \hat{p}_j$, we can rewrite:
\begin{equation}
\nabla_z \ell = -\Big(\sum_{j \neq y} \hat{p}_j\Big) h_{T,y} + \sum_{j \neq y} \hat{p}_j\, h_{T,j} = \sum_{j \neq y} \hat{p}_j\, (h_{T,j} - h_{T,y}).
\end{equation}
This reveals that the gradient is a \emph{non-negative linear combination} of pairwise differences between incorrect-class embeddings and the correct-class embedding, with weights $\hat{p}_j$ summing to $1 - \hat{p}_y \le 1$. Taking the norm:
\begin{equation}
\|\nabla_z \ell\|_2 \le \sum_{j \neq y} \hat{p}_j \|h_{T,j} - h_{T,y}\|_2 \le \max_{j \neq y} \|h_{T,j} - h_{T,y}\|_2 \cdot \underbrace{\sum_{j \neq y} \hat{p}_j}_{\le\, 1}.
\end{equation}
Taking the maximum over all possible true classes $y$:
\begin{equation}
\label{eq:kfeat_tight}
K_{\mathrm{feat}} = \max_{j,k} \|h_{T,j} - h_{T,k}\|_2.
\end{equation}

\paragraph{Remark.} This bound depends on the \emph{relative distances} between token embeddings, not their absolute magnitudes. A uniform shift $H_T \to H_T + c\mathbf{1}^\top$ does not change $K_{\mathrm{feat}}$. A naive Cauchy--Schwarz bound gives $K_{\mathrm{feat}}^{\mathrm{naive}} = \sqrt{2}\|H_T\|_2$, which is substantially looser.

\paragraph{From Gradient Bound to Lipschitz Inequality.}
Since $g_T(\cdot, y)$ is continuously differentiable, the gradient bound $\|\nabla_z g_T(z, y)\|_2 \le K_{\mathrm{feat}}$ implies the Lipschitz inequality stated above by the mean value theorem: for any $z_1, z_2$,
\begin{equation*}
g_T(z_2, y) - g_T(z_1, y) = \int_0^1 \nabla_z g_T\!\big(z_1 + t(z_2 - z_1),\, y\big)^\top (z_2 - z_1)\, dt,
\end{equation*}
and Cauchy--Schwarz gives $|g_T(z_2, y) - g_T(z_1, y)| \le K_{\mathrm{feat}}\, \|z_2 - z_1\|_2$.

\paragraph{Applying the Lipschitz Bound.}
Using the Lipschitz property with $z_1 = \phi_T(x)$ and $z_2 = \phi_P(x) W$:
\begin{equation}
\begin{aligned}
\delta &= \big|\mathcal{R}_T - \mathcal{R}_{P \to T}\big| = \Big|\mathbb{E}\big[g_T(\phi_T(x), y) - g_T(\phi_P(x) W, y)\big]\Big| \\
&\le \mathbb{E}\Big[|g_T(\phi_T(x), y) - g_T(\phi_P(x) W, y)|\Big] \\
&\le K_{\mathrm{feat}} \cdot \mathbb{E}_x\!\big[\|\phi_T(x) - \phi_P(x) W\|_2\big].
\end{aligned}
\end{equation}

Applying Jensen's inequality ($\mathbb{E}[\|X\|] \le \sqrt{\mathbb{E}[\|X\|^2]}$) to connect this to the alignment residual:
\begin{equation}
\label{eq:delta_lipschitz}
\delta \le K_{\mathrm{feat}} \cdot \sqrt{\mathbb{E}_x\!\big[\|\phi_T(x) - \phi_P(x) W\|_2^2\big]} = K_{\mathrm{feat}} \cdot \sqrt{\frac{1}{n}\|Z_T - Z_P W\|_F^2}.
\end{equation}



\subsection{Step 3: Exact Geometric Decomposition}
\label{app:procrustes}

We derive the exact identity linking the alignment residual to scale and shape mismatches. The identity holds for every $W \in \mathcal{O}(d)$; we then specialize to the two cases relevant to this paper.

\paragraph{General Identity.}
For any $W \in \mathcal{O}(d)$, expand the squared Frobenius norm:
\begin{equation}
\begin{aligned}
\|Z_T - Z_P W\|_F^2 &= \operatorname{Tr}\!\big[(Z_T - Z_P W)^\top (Z_T - Z_P W)\big] \\
&= \|Z_T\|_F^2 + \|Z_P W\|_F^2 - 2\operatorname{Tr}(Z_T^\top Z_P W).
\end{aligned}
\end{equation}
Since $W$ is orthogonal, $\|Z_P W\|_F^2 = \operatorname{Tr}(W^\top Z_P^\top Z_P W) = \operatorname{Tr}(Z_P^\top Z_P) = \|Z_P\|_F^2$ (cyclic trace + $W^\top W = I_d$). Therefore:
\begin{equation}
\label{eq:procrustes_expand}
\|Z_T - Z_P W\|_F^2 = \|Z_T\|_F^2 + \|Z_P\|_F^2 - 2\operatorname{Tr}(Z_T^\top Z_P W).
\end{equation}
Recall $\rho_M = \|Z_M\|_F/\sqrt{n}$ (so $\|Z_M\|_F^2/n = \rho_M^2$) and $\Omega_W := \operatorname{Tr}(Z_T^\top Z_P W)/(\|Z_T\|_F\|Z_P\|_F)$ (main-text Eq.~\ref{eq:omega_def}). Dividing Eq.~(\ref{eq:procrustes_expand}) by $n$:
\begin{equation}
\frac{1}{n}\|Z_T - Z_P W\|_F^2 = \rho_T^2 + \rho_P^2 - 2\rho_T\rho_P\, \Omega_W.
\end{equation}
Adding and subtracting $2\rho_T\rho_P$ completes the square:
\begin{equation}
\label{eq:exact_decomposition_general}
\frac{1}{n}\|Z_T - Z_P W\|_F^2 = \underbrace{(\rho_T - \rho_P)^2}_{\text{Scale Mismatch}} + \underbrace{2\rho_T\rho_P(1 - \Omega_W)}_{\text{Shape Mismatch}}.
\end{equation}
This is Proposition~\ref{prop:exact_decomposition} in its general form. \qed

\paragraph{Specialization 1: $W = I$ (identity alignment, main text).}
Setting $W = I$ gives the trace form $\Omega \equiv \Omega_{W=I} = \operatorname{Tr}(Z_T^\top Z_P)/(\|Z_T\|_F \|Z_P\|_F)$, and Eq.~(\ref{eq:exact_decomposition_general}) becomes
\begin{equation}
\frac{1}{n}\|Z_T - Z_P\|_F^2 = (\rho_T - \rho_P)^2 + 2\rho_T\rho_P(1 - \Omega),
\end{equation}
which is the form used throughout the main-text experiments and the regularizer of Sec.~\ref{subsec:shape_reg}.

\paragraph{Specialization 2: $W = W_N$ (Procrustes-optimal, ablation).}
The alignment that minimizes the residual is the classical orthogonal Procrustes problem~\cite{schonemann1966generalized, gower2004procrustes}:
\begin{equation}
W_N := \arg\min_{W \in \mathcal{O}(d)} \|Z_T - Z_P W\|_F^2 = \arg\max_{W \in \mathcal{O}(d)} \operatorname{Tr}(Z_T^\top Z_P W),
\end{equation}
where the second equality follows from Eq.~(\ref{eq:procrustes_expand}) since the first two terms are $W$-independent. Its closed-form SVD solution is well known; we restate it here for completeness. Let $Z_T^\top Z_P = U\Sigma V^\top$ be the SVD with $U, V \in \mathcal{O}(d)$ and $\Sigma = \mathrm{diag}(\sigma_1, \ldots, \sigma_d)$, $\sigma_i \ge 0$. By cyclic trace:
\begin{equation}
\operatorname{Tr}(Z_T^\top Z_P W) = \operatorname{Tr}(\Sigma\, V^\top W U) = \operatorname{Tr}(\Sigma R), \quad R := V^\top W U \in \mathcal{O}(d).
\end{equation}
Since $\sigma_i \ge 0$ and $|R_{ii}| \le 1$ for any orthogonal $R$, $\operatorname{Tr}(\Sigma R) = \sum_i \sigma_i R_{ii}$ is maximized at $R = I_d$, giving:
\begin{equation}
\label{eq:W_optimal}
W_N = V U^\top, \qquad \max_{W} \operatorname{Tr}(Z_T^\top Z_P W) = \sum_i \sigma_i = \|Z_T^\top Z_P\|_*.
\end{equation}
Therefore $\Omega_{W_N} = \Omega_N := \|Z_T^\top Z_P\|_*/(\|Z_T\|_F\|Z_P\|_F)$ (nuclear form, Appendix~\ref{app:tightness} Eq.~\ref{eq:omega_nuclear}), and Eq.~(\ref{eq:exact_decomposition_general}) at $W = W_N$ becomes
\begin{equation}
\min_{W \in \mathcal{O}(d)} \frac{1}{n}\|Z_T - Z_P W\|_F^2 = (\rho_T - \rho_P)^2 + 2\rho_T\rho_P(1 - \Omega_N),
\end{equation}
the form underlying the Sec.~\ref{subsec:ablation} ablation. Note that $W_N$ minimizes only the feature alignment residual $\delta(W)$, not the full bound $\delta(W) + \gamma(W)$, since the head term also depends on $W$ via $W H_T - H_P$. Appendix~\ref{app:joint_opt} treats the joint optimization in full and shows that $W_N$ remains the operative choice in the frozen-head LoRA and FP16-head PTQ settings studied here, where $\gamma$ either vanishes or is $W$-independent.

\paragraph{Assembling the Feature Bound.}
Combining Eq.~(\ref{eq:delta_lipschitz}) with the general decomposition Eq.~(\ref{eq:exact_decomposition_general}):
\begin{equation}
\label{eq:delta_final}
\delta \le K_{\mathrm{feat}} \sqrt{(\rho_T - \rho_P)^2 + 2\rho_T\rho_P(1 - \Omega_W)}.
\end{equation}
At $W = I$, replace $\Omega_W$ with $\Omega$; at $W = W_N$, with $\Omega_N$.



\subsection{Step 4: Bounding the Head Discrepancy $\gamma$}
\label{app:head_bound}

\paragraph{Lipschitz Property with Respect to Logits.}
The head discrepancy measures how much predictions differ when two different ``effective heads'' ($W H_T$ vs.\ $H_P$) operate on the same proxy features. Let $z = \phi_P(x)$ denote the (unaligned) proxy feature. Then:
\begin{equation}
\gamma = \big|\mathcal{R}_{P \to T} - \mathcal{R}_P\big| = \Big|\mathbb{E}\!\big[\ell(z \cdot W H_T, y) - \ell(z \cdot H_P, y)\big]\Big|.
\end{equation}
To bound this, we use the Lipschitz property of cross-entropy with respect to \emph{logits}. The gradient of $\ell$ with respect to $v$ is $\nabla_v \ell = \hat{p} - e_y$, with norm:
\begin{equation}
\|\nabla_v \ell\|_2 = \|\hat{p} - e_y\|_2.
\end{equation}
Since $\hat{p}$ lies on the probability simplex, $\|\hat{p}\|_2^2 \le \|\hat{p}\|_1^2 = 1$, and therefore $\|\hat{p} - e_y\|_2^2 = \|\hat{p}\|_2^2 - 2\hat{p}_y + 1 \le 2 - 2\hat{p}_y \le 2$. The supremum $\sqrt{2}$ is approached as the model's predicted distribution concentrates on an incorrect class ($\hat{p}_y \to 0$). Thus $K_{\mathrm{pred}} \le \sqrt{2}$, yielding the Lipschitz inequality $|\ell(v_1, y) - \ell(v_2, y)| \le K_{\mathrm{pred}}\, \|v_1 - v_2\|_2$ on logits.

\paragraph{Deriving the Second-Moment Projection.}
Define the head misalignment matrix $\Delta H := W H_T - H_P \in \mathbb{R}^{d \times V}$. The logit difference for a single sample is $z \cdot \Delta H = \phi_P(x)(W H_T - H_P)$, and:
\begin{equation}
\begin{aligned}
\gamma &\le \mathbb{E}\!\big[|\ell(z \cdot W H_T, y) - \ell(z \cdot H_P, y)|\big] \\
&\le K_{\mathrm{pred}} \cdot \mathbb{E}_z\!\big[\|z \cdot \Delta H\|_2\big].
\end{aligned}
\end{equation}
Applying Jensen's inequality on the concave square root:
\begin{equation}
\big(\mathbb{E}[\|z \cdot \Delta H\|_2]\big)^2 \le \mathbb{E}\!\big[\|z \cdot \Delta H\|_2^2\big].
\end{equation}
Expanding the squared norm using the trace operator:
\begin{equation}
\begin{aligned}
\mathbb{E}\!\big[\|z \cdot \Delta H\|_2^2\big] &= \mathbb{E}\!\Big[\operatorname{Tr}\!\big(\Delta H^\top z^\top z\, \Delta H\big)\Big] \\
&= \operatorname{Tr}\!\Big(\Delta H^\top \underbrace{\mathbb{E}[z^\top z]}_{\Sigma_P} \Delta H\Big) \\
&= \|\Sigma_P^{1/2} \Delta H\|_F^2.
\end{aligned}
\end{equation}
Taking the square root:
\begin{equation}
\label{eq:gamma_final}
\gamma \le K_{\mathrm{pred}} \cdot \|\Sigma_P^{1/2}(W H_T - H_P)\|_F.
\end{equation}

\paragraph{Interpretation.} The second-moment weighting $\Sigma_P^{1/2}$ projects the head error onto the \emph{active subspace} of the data. If the heads disagree only in directions orthogonal to the feature distribution (the null space of $\Sigma_P$), this disagreement has zero cost---it does not affect any prediction. This is strictly tighter than the spectral bound $\gamma \le K_{\mathrm{pred}} \|W H_T - H_P\|_2 \cdot \mathbb{E}[\|z\|_2]$, which assumes worst-case alignment between features and head error.


\subsection{Step 5: Assembling the Unified Bound}
\label{app:assembly}

For any $W \in \mathcal{O}(d)$, combining the risk decomposition (Eq.~\ref{eq:app_triangle}), the feature bound (Eq.~\ref{eq:delta_final}), and the head bound (Eq.~\ref{eq:gamma_final}):
\begin{equation}
|\mathcal{R}_T - \mathcal{R}_P| \;\le\; \delta + \gamma \;\le\; K_{\mathrm{feat}} \sqrt{(\rho_T - \rho_P)^2 + 2\rho_T\rho_P(1 - \Omega_W)} \;+\; K_{\mathrm{pred}} \cdot \|\Sigma_P^{1/2}(W H_T - H_P)\|_F.
\end{equation}
This is Theorem~\ref{thm:unified_bound} in its general form. The two specializations used in the paper:
\begin{itemize}
    \item \textbf{$W = I$ (identity alignment, main text):} $\Omega_W = \Omega$ (trace form) and $W H_T - H_P = H_T - H_P$, giving the bound used throughout Sec.~\ref{sec:experiments} and the regularizer of Sec.~\ref{subsec:shape_reg}.
    \item \textbf{$W = W_N$ (Procrustes-optimal, ablation):} $\Omega_W = \Omega_N$ (nuclear form, achieves the tightest feature-side specialization), giving the bound underlying the Sec.~\ref{subsec:ablation} ablation.
\end{itemize}
This completes the proof of Theorem~\ref{thm:unified_bound}. \qed



