\section{When shape regularization helps: task-dependence}
\label{app:reg_task_dependence}

TruthfulQA-FT, where shape drift dominates downstream forgetting, illustrates the regularizer's intended use case: trace cuts mean $|\Delta\mathcal{R}|$ from 0.843 (no reg) to 0.681 ($-19\%$), outperforming experience replay (0.764, $-9\%$). BBQ-FT serves as a robustness check in the small-$|\Delta\mathcal{R}|$ regime: averaging the per-benchmark $|\Delta\mathcal{R}|$ values from Table~\ref{tab:reg_compare_llama_bbq} gives means of 0.179 (no reg), 0.195 (trace), and 0.241 (replay)---substantially smaller than TruthfulQA's 0.843---reflecting BBQ's modest baseline shape drift (mean baseline $\Omega = 0.932$ vs $0.906$). Trace outperforms replay; per-benchmark substantive wins (detailed below) persist on shape-driven cells.

\paragraph{Geometric mechanism remains intact.}
On BBQ, the contrast between trace and experience replay is most visible in the geometric signal: trace lifts $\Omega$ from 0.93 to 0.98 (a $73\%$ reduction in $1-\Omega$) while experience replay leaves $\Omega$ flat (0.93 $\to$ 0.93). The regularizer's targeted quantity moves as designed.

\paragraph{Per-benchmark wins on shape-driven cells.}
Trace's largest $|\Delta\mathcal{R}|$ reductions appear where forgetting is substantive: on BBQ-FT (Table~\ref{tab:reg_compare_llama_bbq}), TriviaQA $|\Delta\mathcal{R}|$ drops from 0.143 to 0.017 ($-88\%$) and GSM8K from 0.061 to 0.013 ($-79\%$); on TruthfulQA-FT (Table~\ref{tab:reg_compact_llama_truthfulqa}), SQuAD from 1.337 to 1.054 ($-21\%$) and TriviaQA from 2.583 to 2.124 ($-18\%$). On BBQ-FT ARC and MMLU, the no-reg condition shows substantial shape drift ($\Omega$ drops to 0.88, 0.87) with relatively small $|\Delta\mathcal{R}|$ (0.077, 0.353): the no-reg evidence indicates shape drift does not translate proportionally into $|\Delta\mathcal{R}|$ growth on these cells, so trace's shape restoration lacks a proportional $|\Delta\mathcal{R}|$ target. These cells' shorter answer spans further compress the per-cell CE dynamic range, limiting the observable signal-to-noise of any regularization effect; when absolute $|\Delta\mathcal{R}|$ is small, per-sample CE noise forms a relatively larger fraction of the signal. Rank correlation across the 2$\times$5 grid (Sec.~\ref{subsec:predict}, mean Spearman $\approx 0.83$) provides the scale-invariant calibration of the bound's predictive power.

\paragraph{When to apply: empirical validation of the gating signal.}
Shape regularization is indicated when (i) $\Omega$ drifts substantially during fine-tuning and (ii) downstream forgetting is shape-dominated; PRISM's decomposition (Sec.~\ref{subsec:decompose}) provides both signals \emph{before} regularization is applied. To quantify how well the gating prediction matches the empirical trace effect, we compute two per-setting scalars across the four $(\text{model}, \text{fine-tuning task})$ combinations, both at step $300$ on the same five downstream benchmarks:
\begin{itemize}[leftmargin=*,nosep,topsep=2pt]
    \item \textbf{Baseline shape-drift signal} $1{-}\bar{\Omega} := \tfrac{1}{5}\sum_b (1-\Omega_b^{\lambda=0})$, the across-benchmark mean shape mismatch under no regularization, with per-benchmark $\Omega_b^{\lambda=0}$ read from the no-reg rows of Tables~\ref{tab:reg_compare_llama_truthfulqa}--\ref{tab:reg_compare_qwen_bbq}. Larger $1{-}\bar{\Omega}$ = larger headroom under condition~(i).
    \item \textbf{Relative trace effect} $\Delta|\Delta\mathcal{R}|/|\Delta\mathcal{R}|_0 := (\overline{|\Delta\mathcal{R}|}_{\mathrm{trace}} - \overline{|\Delta\mathcal{R}|}_{\lambda=0})/\overline{|\Delta\mathcal{R}|}_{\lambda=0}$: across-benchmark mean of the trace-vs-no-reg relative change in empirical forgetting at $\lambda{=}1.0$ (negative $=$ trace helps; positive $=$ trace hurts).
\end{itemize}

\begin{table}[h]
\centering
\small
\setlength{\tabcolsep}{8pt}
\caption{Diagnostic-gating prediction vs.\ empirical trace effect across the four $(\text{model}, \text{fine-tuning task})$ settings, ordered by decreasing $1{-}\bar{\Omega}$. Trace yields the largest $|\Delta\mathcal{R}|$ reduction where shape drift is largest (Llama TruthfulQA), is essentially neutral on both Qwen settings where $1{-}\bar{\Omega}$ sits at the noise floor, and shows a partial cell-level exception on Llama BBQ explained in the next paragraph.}
\label{tab:reg_gating}
\begin{tabular}{l c c l}
\toprule
\multicolumn{1}{c}{\textbf{Setting}} & $\boldsymbol{1{-}\bar{\Omega}}$ & $\boldsymbol{\Delta|\Delta\mathcal{R}|/|\Delta\mathcal{R}|_0}$ & \multicolumn{1}{c}{\textbf{Gating verdict}} \\
\midrule
Llama TruthfulQA & $0.0937$ & $-19.2\%$ & shape-driven $\Rightarrow$ apply \\
Llama BBQ        & $0.0678$ & $+8.6\%$  & cell-level mixed (see below) \\
Qwen TruthfulQA  & $0.0091$ & $+2.7\%$  & at noise floor $\Rightarrow$ skip \\
Qwen BBQ         & $0.0011$ & $-0.2\%$  & at noise floor $\Rightarrow$ skip \\
\bottomrule
\end{tabular}
\end{table}

Three of four settings match the gating prediction directly: trace produces the largest benefit where $1{-}\bar{\Omega}$ is largest (Llama TruthfulQA: $-19.2\%$), and is essentially neutral (magnitude $<3\%$) on both Qwen settings where $1{-}\bar{\Omega}$ sits at the noise floor. Llama BBQ is the partial exception---its aggregate $1{-}\bar{\Omega}=0.0678$ exceeds the noise floor (condition~(i) satisfied), but the drift is concentrated on ARC and MMLU where condition~(ii) is violated: substantial shape drift accompanies small $|\Delta\mathcal{R}|$ growth, so trace's shape restoration lacks a proportional $|\Delta\mathcal{R}|$ target (per-benchmark paragraph above). On the two Llama-BBQ benchmarks where condition~(ii) does hold (TriviaQA and GSM8K), trace yields $-88\%$ and $-79\%$ reductions respectively, so the aggregate $+8.6\%$ reflects mixed cell-level behavior rather than a failure of the shape-drift gating itself. Adaptive per-cell deployment of the regularizer---driven by PRISM's per-benchmark decomposition online---is the natural extension and is left to future work, alongside the short-answer fine-tuning regimes noted earlier in this appendix.

\paragraph{Empirical confirmation: Qwen3-8B replication.}
The Qwen3-8B replication (Tables~\ref{tab:reg_compare_qwen_truthfulqa}--\ref{tab:reg_compare_qwen_bbq}, Fig.~\ref{fig:shape_reg_combined_qwen}) extends the regime-dependence story from \emph{tasks} (TruthfulQA-FT vs.\ BBQ-FT on Llama) to \emph{models}, providing direct empirical evidence for condition~(i) above. Qwen3-8B is markedly more robust to LoRA forgetting on these two fine-tuning sources than Llama-3.1-8B: across-benchmark mean $|\Delta\mathcal{R}|$ at $\lambda{=}0$ is $0.263$ on Qwen TruthfulQA-FT vs.\ Llama's $0.843$ ($3.2\times$ smaller), and $0.112$ on Qwen BBQ-FT vs.\ Llama's $0.179$. Baseline shape preservation is correspondingly closer to its ceiling on Qwen ($\Omega$ at $\lambda{=}0$: $0.991$ on Qwen TruthfulQA-FT, $0.999$ on Qwen BBQ-FT vs.\ Llama's $0.906$ and $0.932$), leaving the shape regularizer with essentially no drift to repair: $1-\Omega$ collapses from Llama's $0.094$/$0.068$ to Qwen's $0.009$/$0.001$, an order of magnitude smaller. The differences across no-reg, replay, and trace correspondingly shrink to the evaluation-noise floor: across-benchmark mean $|\Delta\mathcal{R}|$ on Qwen TruthfulQA-FT is $0.263$ / $0.282$ / $0.270$ (no-reg / replay / trace), and on Qwen BBQ-FT is $0.112$ / $0.110$ / $0.112$, with method-to-method spread under $0.02$ on both. This is the same low-SNR phenomenon documented for GSM8K under PTQ in Appendix~\ref{app:per_model_tables} (Table~\ref{tab:gsm8k_outlier}), where the bound's per-benchmark mean Spearman drops from $\ge 0.77$ on ARC/MMLU/SQuAD/TriviaQA to $\approx 0.41$ on GSM8K precisely because $|\Delta\mathcal{R}|$ collapses to $\approx 0.019$. We report the Qwen3-8B regularization numbers in full for transparency and reproducibility, but emphasize that the shape regularizer's mechanism---contracting $1-\Omega$ at the source of LoRA backbone drift---can demonstrate a measurable benefit only when $1-\Omega$ has substantive headroom to contract. Qwen3-8B fine-tuning on TruthfulQA/BBQ violates condition~(i) above, and the Qwen3-8B numbers therefore do not contradict the Llama TruthfulQA-FT result; they confirm the regime-dependence the bound itself predicts.
